---
title: "Where Bobcats Are, and Why"
subtitle: "Question 2 — Bobcat"
---

Bobcats are the easy species to study here, and that ease is itself the problem.
Records are abundant, so the question is not whether there is enough data — it is
how much of the apparent pattern is bobcats and how much is people with cameras
and phones. The occupancy model is built to separate the two.

## Occupancy, not just sightings

A raw map of bobcat sightings shows where people looked as much as where bobcats
are. An occupancy model pulls those apart. It treats each open-space unit as a
site visited repeatedly, models the chance of *detecting* a bobcat given that it
is present, and from that estimates the chance the unit is *occupied* at all. The
detection part absorbs the effort; the occupancy part is what's left.

The fit is a single-species occupancy model over 845 open-space units, built in a
fixed two-stage order: the detection sub-model first, then occupancy with the
detection structure held fixed. The covariate sets and the selection rule were
written down before any fitted number was looked at (Decision 31), so the result
is a test of a pre-registered model, not a search for one that fit.

## Detection is effort, plainly

The detection result is blunt. The single covariate is a per-visit record-count
effort proxy, and it dominates: more recording effort, far higher detection, with
a strength that leaves no room for doubt. Against a no-effort null the effort
model is better by an AICc gap of 408 — not a subtle correction, the main signal
in the detection process.

This is the whole reason the occupancy model exists. A sightings map would read
that effort as bobcat distribution. The model reads it as effort and sets it
aside, so the occupancy surface underneath is not just a map of where people were
looking.

## What explains occupancy

With effort handled, three landscape signals shape where bobcats are. The pattern
needs all three parts together — terrain, land cover, and human footprint. The
model that keeps only one of them falls far behind: a human-footprint-only or a
land-cover-only model sits more than 40 AICc units below the full model, while the
full model and its near-twin carry 85% of the weight between them.

```{r}
#| label: fig-bobc-coef
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Bobcat occupancy coefficients (logit scale, top model m_full), with 95% confidence intervals. Points right of the zero line raise the modelled chance a unit is occupied; points left lower it. Elevation is the strongest signal (positive); human modification (gHM) is the clearest negative. Tree cover leans positive but its interval touches zero. Terrain aspect, slope, grass, and housing density do not separate from zero. Source: Decision 31 occupancy fit (tbl_13)."
#| fig-alt: "A coefficient plot with nine occupancy terms on the vertical axis and effect size on the logit horizontal axis, a dashed line at zero. Mean elevation sits well to the right at about +1.4 with an interval clear of zero. Human modification sits to the left at about -0.56 with an interval clear of zero. Tree cover is positive at about +0.62 with an interval just touching zero. Grass, northerly and easterly aspect, slope, and housing density all sit near zero with intervals crossing it."

library(ggplot2)

# Occupancy (psi) coefficients from the top model m_full (Decision 31 fit).
# Estimates and SEs inlined from summary(mod); CI = estimate +/- 1.96*SE.
co <- data.frame(
  term = c("Mean elevation", "Slope", "Aspect (north)", "Aspect (east)",
           "Tree cover", "Grass cover", "Human modification (gHM)",
           "Housing density (log)"),
  est  = c( 1.4011, 0.0896, 0.1561, -0.0860, 0.6166, 0.4125, -0.5608, 0.1691),
  se   = c( 0.488,  0.275,  0.186,   0.168,  0.326,  0.272,   0.204,  0.168)
)
co$lo <- co$est - 1.96 * co$se
co$hi <- co$est + 1.96 * co$se

# Significance at the conventional 0.05 level drives colour, not the story —
# the caption states the directions in words.
co$sig <- ifelse(co$lo > 0 | co$hi < 0, "clear", "unresolved")

# Order by effect size for readability.
co$term <- factor(co$term, levels = co$term[order(co$est)])

col_pos <- "#2f6b3a"  # forest green — raises occupancy
col_neg <- "#9a4a2f"  # rust — lowers occupancy
col_nul <- "#8a8f9a"  # grey — interval crosses zero

co$col <- ifelse(co$sig == "unresolved", col_nul,
                 ifelse(co$est > 0, col_pos, col_neg))

ggplot(co, aes(x = est, y = term, color = col)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#b9b4a8") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.25, linewidth = 0.6) +
  geom_point(size = 2.6) +
  scale_color_identity() +
  labs(
    x = "Effect on occupancy (logit scale)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "#edeae3"),
    axis.text  = element_text(color = "#1a1a18"),
    axis.title = element_text(color = "#1a1a18"),
    plot.margin = margin(8, 14, 8, 8)
  )
```

**Elevation is the strongest signal, and it is positive.** Bobcat occupancy rises
with mean unit elevation — the higher, less-developed units of the Coast Ranges
and Diablo are the strongholds. **Human modification is the clearest negative.**
As the gHM index rises, modelled occupancy falls: this is the coexistence limit
showing up directly in the numbers. **Tree cover leans positive** but its interval
just touches zero, so it reads as a lean, not a firm effect. Grass cover, slope,
terrain aspect, and housing density do not separate from zero in this fit.

The housing result is worth one note. Housing density enters positive but
unresolved, while gHM enters clearly negative. The two are correlated, and the
project deliberately lets gHM carry the human-pressure axis (Decision 23). Read
together, the human signal in bobcat occupancy is real and negative — it just
lives in the gHM term, not the housing one.

### The occupancy surface

The map below is the model-corrected result: predicted occupancy for each
open-space unit, after detection effort has been modelled out. It is the layer to
trust for distribution — not the raw sightings, and not the kernel density that
follows.

```{r}
#| label: fig-bobc-psi
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Predicted bobcat occupancy (ψ) by open-space unit, from the model-averaged occupancy fit. Darker green is higher predicted occupancy. Grey units are not modelled — they lack covariate coverage, so no prediction exists; grey means no estimate, not low occupancy. The surface covers the 845 fitted units; the 284 grey units carry geometry only. Detection effort has been modelled out of this surface, which is what separates it from the raw record maps below. Source: model-averaged occu fit (m_full + m_fullgrad)."

library(sf)
library(leaflet)
library(dplyr)

# Rendered from site/, so data lives one level up at ../data/. A page-relative
# path is predictable here; here::here() can mis-anchor to site/ because Quarto
# plants its own project marker.
psi <- st_read("../data/processed/occu_bobc_pred_unit_3310.gpkg", quiet = TRUE) |>
  st_transform(4326)

modelled  <- psi[!is.na(psi$psi_pred), ]
notmodel  <- psi[ is.na(psi$psi_pred), ]

pal <- colorNumeric("Greens", domain = c(0, 1), na.color = "#cfccc4")

leaflet(options = leafletOptions(minZoom = 7)) |>
  addProviderTiles("CartoDB.Positron") |>
  # not-modelled units first, underneath, in flat grey
  addPolygons(
    data = notmodel,
    fillColor = "#cfccc4", fillOpacity = 0.55,
    color = "#9a978f", weight = 0.5, opacity = 0.8,
    label = ~paste0("Unit ", unit_id, " — not modelled (no covariate coverage)")
  ) |>
  # modelled units on top, green scale
  addPolygons(
    data = modelled,
    fillColor = ~pal(psi_pred), fillOpacity = 0.8,
    color = "#4a4a48", weight = 0.4, opacity = 0.7,
    highlightOptions = highlightOptions(weight = 2, color = "#222", bringToFront = TRUE),
    label = ~sprintf("Unit %s — \u03c8 = %.2f (SE %.2f)", unit_id, psi_pred, psi_se)
  ) |>
  addLegend(
    position = "bottomright", pal = pal, values = c(0, 1),
    title = "Predicted occupancy (\u03c8)", opacity = 0.8
  ) |>
  addControl(
    "<div style='font-size:11px;line-height:1.3;max-width:200px'>
     Grey = not modelled (no covariate coverage), <b>not</b> low occupancy.
     Effort is modelled out of this surface.</div>",
    position = "topright"
  )
```

This surface is where the effort correction pays off. A raw sightings map would
put the darkest signal wherever recording is densest — near trailheads, near
cities. The occupancy model moves the signal to where the *landscape* supports
bobcats: up in elevation, away from human modification. Where the two disagree is
the effort bias made visible, and it is the heart of Question 5.

## Two more views, and why they need care

The occupancy surface is the model-corrected view. Two descriptive layers sit
alongside it, and both carry the effort problem the occupancy model was built to
escape — so both are read with that caveat attached.

**Kernel density.** A smoothed surface of where bobcat records fall. The smoothing
scale — the bandwidth — is set by a pre-registered rule, not by hand and not by
the data-driven default, because the default collapses toward a tiny bandwidth
that just maps the densest clusters of *effort* (Decision 28). Even with the rule,
a kernel-density surface of opportunistic records maps detection effort as much as
bobcat density. It is shown as a descriptive layer, not as an abundance estimate.

**Hot spots.** A Getis-Ord Gi\* analysis flags units with significantly high or low
record counts relative to their neighbours. The honest problem, carried over from
the tiger project: a raw-count hot spot is partly an observer-effort hot spot. So
each hot unit is cross-read against the effort surface and labelled — TRUSTED
where the cluster is not explained by effort, EFFORT-FLAGGED where it is. The map
shows the label, not just the count.

## The detection-bias caveat, stated once and carried

Every descriptive map in this section — kernel density, hot spots, even the raw
record counts behind them — reflects where people recorded bobcats, not only where
bobcats are. That is not a footnote to bury; it is the point of Question 5, and it
is why the occupancy surface, which models effort explicitly, is the layer to
trust for distribution. The two other layers are shown because they are
informative about pattern, with the effort caveat attached to each, not because
they are unbiased.

This is also the cleanest contrast in the whole project. The bobcat occupancy
model *corrects* for effort and, where it does, its picture diverges from the raw
kernel density. The puma corridors in the next section do not depend on effort at
all — they are built from landscape resistance, not from sightings — so they
converge with the density surface for a different reason. Same region, two species,
two entirely different relationships between what was recorded and what is there.
