---
title: "The Protected-Land Fabric"
subtitle: "Question 1 — Landscape"
---

The first question is about the ground itself, before any animal moves across it.
How much protected open space is there in the ten-county Bay Area, how is it held,
and how does it break apart when you look at it the way a wide-ranging carnivore
would — as a set of patches large enough to hold a home range, or not?

## Two kinds of protection

Protected land here comes in two forms, and the difference matters for a puma.
Fee land is owned outright for conservation: the parks, watershed lands, and
open-space preserves that most people picture. Easement land is private
ground — often working ranchland — where the owner has sold or donated the right
to develop it, so it stays open. To a puma crossing it, an easement grazing
parcel next to a fee preserve is one continuous piece of habitat. Tenure is a
legal distinction, not an ecological one.

The project keeps both. The connectivity track uses a union of the two sources:
CPAD fee land and CCED conservation easements, merged into one protected-land
layer, with fee taking precedence wherever the two map the same ground so no area
is counted twice.

## How much, and how much is new

The union holds **5,996.5 km²** of protected land across the ten counties. Fee
land accounts for **4,720.8 km²** of that, spread over 1,129 CPAD units. The
easements add **1,275.7 km²** of genuinely new protected ground — a roughly 27%
increase over the fee footprint alone — after removing 498.2 km² where easements
overlapped land already protected in fee.

That 27% is the point worth holding on to. Conservation easements are not a
rounding error on the map of Bay Area protected land. More than a quarter of the
connectivity fabric — much of it working ranchland along the inner Coast Ranges —
exists only because of easements. Leave them out and the protected-land picture
looks substantially more fragmented than it is.

One limit on this comes with the easement data. Where the easement layer has no
record, that does not mean the land is unprotected — only that no easement is
mapped there. The union labels tenure where the data exists; it does not fill
gaps. Read the easement totals as a floor, not a full accounting.

## From protected land to habitat patches

A wide-ranging carnivore does not experience protected land as a legal inventory.
It experiences it as patches — connected pieces of open ground it can settle in
or move through. So the next step dissolves the union: fee and easement melt
together wherever they touch, and the result is cut into distinct patches by
physical contiguity, not by ownership.

Most of those patches are small. The patch-area distribution is a smooth size
continuum, from thousands of tiny fragments up to a handful of very large blocks,
with no natural break to cut on. A break would let the data choose the threshold;
there isn't one, so the threshold comes from the animal instead.

A puma home range spans tens to hundreds of square kilometres. A protected patch
smaller than about 5 km² cannot hold a meaningful fraction of even one such range,
so it works as a stepping-stone at most, not as a place a resident animal lives.
Setting that 5 km² floor — an ecological scale, not a number read off a curve —
leaves **164 core patches** large enough to anchor a home range. The floor still
retains 82.5% of all protected area, because the area that gets dropped sits in
thousands of tiny fragments that together hold very little ground.

The map below shows the whole fabric in three layers: the fee-owned CPAD land and
the CCED easements underneath (each toggleable), and the 164 core patches drawn
prominently on top. Hover any core patch for its name and area. Toggle the two
tenure layers to see how much of the connectivity fabric — much of it inner-Coast-Range
ranchland — exists only because of easements.

```{r}
#| label: fig-protected-fabric
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The ten-county protected-land fabric in three layers. Fee-owned CPAD land (green) and CCED conservation easements (gold) form the protected base; toggle each with the control at top-right. The 164 core patches — protected blocks of at least 5 km², large enough to anchor a puma home range — are drawn in dark outline on top; hover any one for its name and area. The two named corridor endpoints (Santa Cruz Mountains, southern Diablo Range) are called out. Boundaries only; no occurrence data."

library(sf)
library(leaflet)
library(htmltools)

# page-relative paths (NEVER here::here)
union_path <- "../data/interim/protected_union_bayarea_3310.gpkg"
core_path  <- "../data/processed/lcp_puma_core_patches_3310.gpkg"

union_sf <- sf::st_read(union_path, quiet = TRUE)
core_sf  <- sf::st_read(core_path,  quiet = TRUE)

# split the union by tenure (single layer, attribute-split)
fee_sf      <- union_sf[union_sf$protection_type == "fee", ]
easement_sf <- union_sf[union_sf$protection_type == "easement", ]

fee_wgs      <- sf::st_transform(fee_sf, 4326)
easement_wgs <- sf::st_transform(easement_sf, 4326)
core_wgs     <- sf::st_transform(core_sf, 4326)

# core hover label: named range where it exists, else "Core patch" + area
core_wgs$lab <- ifelse(
  !is.na(core_wgs$range_name),
  sprintf("<b>%s</b><br>%.0f km\u00b2 (corridor endpoint)",
          core_wgs$range_name, core_wgs$area_km2),
  sprintf("Core patch<br>%.1f km\u00b2", core_wgs$area_km2)
)
core_lab <- lapply(core_wgs$lab, htmltools::HTML)

# named endpoints, for a bolder outline on top
core_named <- core_wgs[!is.na(core_wgs$range_name), ]

col_fee  <- "#2e7d5b"  # green — fee land
col_ease <- "#c8922b"  # gold  — easement
col_core <- "#3b2f2f"  # dark ink — core patch outline

legend_html <- htmltools::HTML(sprintf(
  "<div style='background:#fff;padding:8px 10px;border-radius:4px;
               box-shadow:0 1px 4px rgba(0,0,0,0.2);font-size:12px;
               line-height:1.5;text-align:left;'>
     <div style='font-weight:600;margin-bottom:5px;'>Protected-land fabric</div>
     <div style='display:flex;align-items:center;gap:7px;'>
       <span style='width:15px;height:13px;background:%s;opacity:0.55;
         border:1px solid %s;display:inline-block;flex:none;'></span>Fee (CPAD)</div>
     <div style='display:flex;align-items:center;gap:7px;'>
       <span style='width:15px;height:13px;background:%s;opacity:0.55;
         border:1px solid %s;display:inline-block;flex:none;'></span>Easement (CCED)</div>
     <div style='display:flex;align-items:center;gap:7px;'>
       <span style='width:15px;height:13px;background:rgba(0,0,0,0);
         border:2px solid %s;display:inline-block;flex:none;'></span>Core patch (&ge; 5 km\u00b2)</div>
   </div>",
  col_fee, col_fee, col_ease, col_ease, col_core
))

leaflet(options = leafletOptions(minZoom = 8)) |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}"),
    attribution = "Tiles &copy; Esri") |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Reference/MapServer/tile/{z}/{y}/{x}")) |>
  # fee land (toggle)
  addPolygons(data = fee_wgs, fillColor = col_fee, fillOpacity = 0.45,
              stroke = TRUE, color = col_fee, weight = 0.3, opacity = 0.6,
              group = "Fee (CPAD)") |>
  # easement land (toggle)
  addPolygons(data = easement_wgs, fillColor = col_ease, fillOpacity = 0.5,
              stroke = TRUE, color = col_ease, weight = 0.3, opacity = 0.6,
              group = "Easement (CCED)") |>
  # 164 core patches, prominent, one colour, hover labels
  addPolygons(data = core_wgs, fill = FALSE,
              color = col_core, weight = 1.4, opacity = 0.9,
              label = core_lab,
              highlightOptions = highlightOptions(weight = 3, color = "#000",
                                                  bringToFront = TRUE),
              group = "Core patches (164)") |>
  # named endpoints, bolder outline on top
  addPolygons(data = core_named, fill = FALSE,
              color = "#111", weight = 2.6, opacity = 1,
              group = "Core patches (164)") |>
  addLayersControl(
    overlayGroups = c("Fee (CPAD)", "Easement (CCED)", "Core patches (164)"),
    options = layersControlOptions(collapsed = FALSE),
    position = "topright") |>
  addControl(html = legend_html, position = "bottomleft") |>
  # fitBounds MUST be last
  fitBounds(-123.53354, 36.85065, -121.20830, 38.86399)
```

Turn the easement layer off and the map opens up with gaps along the inner Coast
Ranges — that missing ground is the 27% of the fabric that easements carry. Turn it
back on and those gaps close. The two named cores, the Santa Cruz Mountains and the
southern Diablo Range, are the endpoints the puma connectivity section links.

```{r}
#| label: fig-floor-scan
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Patch count against the minimum-area floor. Each candidate floor is plotted with the number of patches that survive it; the amber point marks the chosen 5 km² floor, which keeps 164 patches and 82.5% of all protected area. The curve is a smooth continuum with no natural break, so the floor is set by the puma home-range scale rather than by a gap in the data. Source: Decision 32 floor scan (tbl_17)."
#| fig-alt: "A line chart on a log vertical axis showing the number of protected patches surviving each minimum-area floor from 0.01 to 10 square kilometres. The count falls smoothly from about 2,356 patches to 86 with no sharp break. The 5 square kilometre floor is marked with a point and label reading 164 patches, 82.5 percent area retained."

library(ggplot2)

# Decision 32 floor scan (tbl_17) — inlined; no disk read, no geometry.
floor_scan <- data.frame(
  floor_km2      = c(0.01, 0.10, 0.50, 1.00, 2.00, 5.00, 10.00),
  patches_kept   = c(2356, 1429, 696, 464, 307, 164, 86),
  area_retained  = c(99.9, 99.3, 96.5, 93.7, 90.0, 82.5, 73.7)
)

col_line   <- "#5a6070"  # slate
col_choice <- "#b07d20"  # amber — the chosen floor
col_ink    <- "#1a1a18"

chosen <- floor_scan[floor_scan$floor_km2 == 5.00, ]

ggplot(floor_scan, aes(x = floor_km2, y = patches_kept)) +
  geom_line(color = col_line, linewidth = 0.7) +
  geom_point(color = col_line, size = 2) +
  geom_point(data = chosen, color = col_choice, size = 4) +
  scale_x_log10(
    breaks = floor_scan$floor_km2,
    labels = c("0.01", "0.10", "0.50", "1", "2", "5", "10")
  ) +
  scale_y_log10(
    breaks = c(100, 200, 500, 1000, 2000),
    labels = scales::comma
  ) +
  labs(
    x = "Minimum-patch-area floor (km², log scale)",
    y = "Patches retained (log scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#edeae3"),
    axis.title = element_text(color = col_ink),
    axis.text  = element_text(color = col_ink),
    plot.margin = margin(8, 14, 8, 8)
  )
```

This is the floor scan, not the raw size distribution: it shows how many patches
survive each candidate threshold, and it is the evidence that the 5 km² cut is not
severe. Most of what the floor removes is thousands of tiny fragments; the
protected area they hold barely moves the total.

This is the structural shape of the problem. Of nearly 4,000 protected patches,
only 164 are large enough to anchor a wide-ranging carnivore, and the habitat
value concentrates in a small number of very large cores — the Santa Cruz
Mountains block at 177 km², the southern Diablo Range at 500 km², and a short list
of others. The rest of the protected fabric is stepping-stones and fragments
between them.

## What this section does and does not claim

This is a description of the protected land, built from protected-area boundaries
alone. It uses no occurrence records, so the detection-effort caveat that governs
the occurrence maps later in this project does not apply here — there is nothing
in these numbers that reflects where people looked for animals. The patches are
where the protected ground is, not where the cats were seen.

What it does not tell you is whether an animal can actually move between those 164
cores. A patch being large is not the same as a patch being reachable. Two cores
can sit a few kilometres apart on the map and be effectively cut off from each
other by a freeway, a reservoir, or the developed bayshore. Turning this static
fabric into a question about movement — which cores connect, and where the
connections are weakest — is the work of the puma connectivity section (Question
3).
