---
title: "Room to Move"
subtitle: "Question 3 — Puma connectivity"
---

The puma question is the opposite of the bobcat one. There are too few puma
records to model where the animals are, and the precise locations that do exist
are sensitive enough that publishing them would do harm. So this section does not
map where pumas were seen. It maps where the landscape would let one move — built
from the shape of the land, not from sightings.

## Built from resistance, not sightings

Every layer in this section rests on a resistance surface: a map of how costly
each 1 km cell is for a puma to cross, high where the land is developed, steep, or
busy with roads, low where it is protected open ground. The surface is assembled
from land cover, terrain, human modification, and traffic — not from where anyone
recorded a puma.

That has one consequence worth stating up front, because it is the mirror image of
the bobcat result. The bobcat occupancy model had to *correct* for recording
effort, because it was built from sightings. The puma corridors carry no effort
term at all — there are no sightings in them to be biased. Where the puma corridors
agree with a density surface, they agree for a structural reason, not because both
were shaped by the same observers. This is the clean half of Question 5.

## Why you see a band, not a line

The core result is a corridor between the two largest habitat blocks in the
region: the Santa Cruz Mountains and the southern Diablo Range, linked across the
south Bay through the Coyote Valley pinch where US-101 cuts the gap. The
least-cost path between them runs 37.2 km and passes within 1.6 km of the verified
Coyote Valley crossing.

But the maps below lead with a **band**, not that line, and the reason is a real
limit, not a stylistic choice. A least-cost path is a single-cell-wide thread
through a 1 km grid. It looks precise, and that precision is false: move the input
weights a little and the thread shifts by a kilometre or more. The honest product
is the **swath** — the zone of low-cost cells around the path — shown in two tiers:

- a **core** band (the lowest-cost 2% of corridor cells, 409 km²) — high-confidence corridor,
- a **context** band (the lowest-cost 5%, 1,021 km²) — the permeable flanks around it.

The swath says what the analysis can actually support: *the corridor runs through
here, in this zone*, not *the corridor is this exact line*. The precise pinch at
Coyote Valley is located separately, by road geometry in the crossing analysis
below — not by the swath, whose 1 km grain is too coarse to resolve a
sub-kilometre constriction.

```{r}
#| label: fig-puma-swath
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The Santa Cruz Mountains ↔ southern Diablo Range puma corridor, shown as a two-tier swath over the protected core habitat it connects. The darker core band is the lowest-cost 2% of corridor cells (409 km²), a high-confidence corridor; the lighter context band is the lowest-cost 5% (1,021 km²), the permeable flanks around it. Faint green blocks are the 164 core habitat patches; the two bold-outlined blocks are the corridor endpoints — Santa Cruz Mountains and the southern Diablo Range. The corridor runs through this zone, not along a single line — the band is the honest product at a 1 km grain. The precise US-101 pinch at Coyote Valley is located by road geometry in the crossing map below, not by this swath. No puma locations are shown; this is generalised corridor geometry. Source: least-cost corridor (Decision 33) over core patches (Decision 32)."

library(sf)
library(leaflet)

# Rendered from site/; data one level up. Page-relative path (here::here can
# mis-anchor under Quarto).
sw <- st_read("../data/processed/lcp_puma_scmtns_to_diablo_swath_3310.gpkg",
              quiet = TRUE) |>
  st_transform(4326)

# Core habitat patches for context — the protected blocks the corridor connects.
# Endpoints are flagged in the data (linkage_role); no hardcoding.
cp <- st_read("../data/processed/lcp_puma_core_patches_3310.gpkg", quiet = TRUE) |>
  st_transform(4326)
cp_plain    <- cp[is.na(cp$linkage_role), ]
cp_endpoint <- cp[!is.na(cp$linkage_role), ]

context_band <- sw[sw$tier == "context", ]
core_band    <- sw[sw$tier == "core", ]

col_core    <- "#7d3f16"  # deep amber-brown — high-confidence corridor
col_context <- "#d8a25e"  # light amber — permeable flanks

leaflet(options = leafletOptions(minZoom = 7)) |>
  addTiles(
    urlTemplate = "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}",
    attribution = "Tiles &copy; Esri"
  ) |>
  addTiles(
    urlTemplate = "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Reference/MapServer/tile/{z}/{y}/{x}"
  ) |>
  # core habitat patches — faint context underneath everything
  addPolygons(
    data = cp_plain,
    fillColor = "#5a7d5a", fillOpacity = 0.22,
    color = "#5a7d5a", weight = 0.4, opacity = 0.4,
    label = ~sprintf("Core habitat patch %s (%.0f km²)", patch_id, area_km2)
  ) |>
  # the two corridor endpoints — same faint fill, bold border, named
  addPolygons(
    data = cp_endpoint,
    fillColor = "#5a7d5a", fillOpacity = 0.28,
    color = "#274427", weight = 2.0, opacity = 0.9,
    label = ~sprintf("%s — corridor endpoint (%.0f km²)", range_name, area_km2)
  ) |>
  # context flanks underneath
  addPolygons(
    data = context_band,
    fillColor = col_context, fillOpacity = 0.45,
    color = col_context, weight = 0.5, opacity = 0.7,
    label = "Context band — lowest-cost 5% (1,021 km²), permeable flanks"
  ) |>
  # core corridor on top
  addPolygons(
    data = core_band,
    fillColor = col_core, fillOpacity = 0.7,
    color = "#4a2409", weight = 0.6, opacity = 0.8,
    label = "Core band — lowest-cost 2% (409 km²), high-confidence corridor"
  ) |>
  addControl(
    position = "bottomleft",
    html = htmltools::HTML(sprintf(
      "<div style='background:#fff;padding:8px 10px;border-radius:4px;
                   box-shadow:0 1px 4px rgba(0,0,0,0.2);font-size:12px;
                   line-height:1.5;text-align:left;max-width:200px;'>
         <div style='font-weight:600;margin-bottom:5px;'>Puma corridor swath</div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:15px;height:13px;background:%s;border:1px solid #4a2409;
                        display:inline-block;flex:none;'></span>Core (409 km²)</div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:15px;height:13px;background:%s;border:1px solid #b98545;
                        display:inline-block;flex:none;'></span>Context (1,021 km²)</div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:15px;height:13px;background:#5a7d5a55;border:2px solid #274427;
                        display:inline-block;flex:none;'></span>Corridor endpoints</div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:15px;height:13px;background:#5a7d5a55;border:1px solid #5a7d5a;
                        display:inline-block;flex:none;'></span>Other core habitat</div>
         <div style='margin-top:5px;color:#5a6070;font-size:11px;'>
           Corridor runs through this zone, not a single line. No puma locations shown.</div>
       </div>",
      col_core, col_context
    ))
  ) |>
  fitBounds(-123.53354, 36.85065, -121.20830, 38.86399)
```

## Where the corridor meets the roads

A corridor is only as good as its worst crossing. Where the swath meets a major
road, the road is a barrier whose severity scales with traffic, not just with
presence — a quiet two-lane road and a freeway are not the same obstacle. So each
crossing is ranked by traffic volume.

The top barrier is **US-101 at Coyote Valley, at 142,000 vehicles a day.** Getting
that number right took a correction. An earlier version of the analysis, using a
spatial traffic-data join, dropped US-101 here to a modelled floor of about 80,000
— a 43% underestimate — and let a lower-traffic arterial outrank it. The fix
referenced the traffic data along the highway route by postmile, recovered the
real 142k, and re-ran the corridor (Decision 34). US-101 is now correctly the
top-ranked crossing.

One honest detail in that ranking: crossings are ranked *within* traffic-data
confidence tiers, not on the raw number alone. A road whose volume is measured
directly outranks one whose volume is filled in by name-matching, even when the
filled-in number is close. This stops a lower-confidence estimate from
masquerading as the top barrier — the exact failure the US-101 correction fixed.

```{r}
#| label: fig-puma-crossings
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Where the puma corridor crosses roads, ranked by traffic, zoomed to the crossing zone in the south Bay. All 290 crossing segments are shown on a neutral road basemap so each sits on its road; line thickness scales with traffic volume (AADT) and solid lines are directly-measured volumes while dashed lines are estimated (modelled, name-filled, or spatially filled — read these as lower-confidence). The four named roads the corridor crosses are outlined and labelled; US-101 at Coyote Valley, at 142,000 vehicles a day, is the top barrier by a wide margin. Unnamed high-traffic links (freeway ramps) appear as thick base-layer lines but are not promoted to named barriers, since one class label covers many separate crossings. No puma locations are shown. Source: corridor × road crossings, AADT by Decision 34."

library(sf)
library(leaflet)
library(dplyr)

cr <- st_read("../data/processed/lcp_puma_scmtns_to_diablo_crossings_3310.gpkg",
              quiet = TRUE) |>
  st_transform(4326)

# Core swath band as faint background context — the corridor these roads cross.
core_ctx <- st_read("../data/processed/lcp_puma_scmtns_to_diablo_swath_3310.gpkg",
                    quiet = TRUE) |>
  st_transform(4326)
core_ctx <- core_ctx[core_ctx$tier == "core", ]

# Confidence: measured volumes solid, estimated volumes dashed.
measured_tiers <- c("measured", "measured_route_pm")
cr$is_measured <- cr$aadt_source %in% measured_tiers

# Line weight scales with traffic (sqrt so the 142k road doesn't dwarf everything).
cr$wt <- 1.2 + 4.5 * sqrt(cr$aadt / max(cr$aadt, na.rm = TRUE))

# Colour ramp by AADT — cool blue (quiet) to magenta (busy barrier), chosen to
# stand clear of the warm orange/red road colours on the basemap.
pal <- colorNumeric(c("#3b8bc2", "#5e4fa2", "#8c2d8f", "#c31d6b"),
                    domain = c(0, max(cr$aadt, na.rm = TRUE)))

# Named roads to highlight (real road identities only; parenthetical class
# labels like "(trunk_link)" lump many separate crossings, so they stay base-only).
named_roads <- c("route 101 (motorway)", "Santa Cruz Highway",
                 "route 17 (motorway)", "Saratoga-Los Gatos Road")
hi <- cr[cr$road_label %in% named_roads, ]

leaflet(options = leafletOptions(minZoom = 7)) |>
  addTiles(
    urlTemplate = "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}",
    attribution = "Tiles &copy; Esri"
  ) |>
  # road overlay: grey roads + labels on the neutral base, no relief/colour
  addTiles(
    urlTemplate = "https://services.arcgisonline.com/arcgis/rest/services/Reference/World_Transportation/MapServer/tile/{z}/{y}/{x}",
    attribution = "Roads &copy; Esri, HERE, Garmin, &copy; OpenStreetMap contributors"
  ) |>
  # core corridor band as background — the corridor the roads cross
  addPolygons(
    data = core_ctx,
    fillColor = "#7d3f16", fillOpacity = 0.20,
    color = "#7d3f16", weight = 1.0, opacity = 0.5,
    label = "Core corridor band (the corridor these roads cross)"
  ) |>
  # all 290 crossing segments — base layer, thickness by AADT, dash by confidence
  addPolylines(
    data = cr,
    color = ~pal(aadt), weight = ~wt + 1, opacity = 0.95,
    dashArray = ~ifelse(is_measured, "", "5 5"),
    label = ~sprintf("%s — %s AADT (%s)", road_label,
                     format(round(aadt), big.mark = ","), aadt_source)
  ) |>
  # named top roads — dark casing on top so they read as the ranked barriers
  addPolylines(
    data = hi,
    color = "#111111", weight = 2, opacity = 0.9, fill = FALSE,
    label = ~sprintf("%s — rank %d, %s AADT (%s)", road_label, rank_overall,
                     format(round(aadt), big.mark = ","), aadt_source),
    popup = ~sprintf("<b>%s</b><br>Rank %d of 290<br>%s vehicles/day<br>Source: %s",
                     road_label, rank_overall,
                     format(round(aadt), big.mark = ","), aadt_source)
  ) |>
  addControl(
    position = "bottomleft",
    html = htmltools::HTML(
      "<div style='background:#fff;padding:8px 10px;border-radius:4px;
                   box-shadow:0 1px 4px rgba(0,0,0,0.2);font-size:12px;
                   line-height:1.4;text-align:left;max-width:210px;'>
         <div style='font-weight:600;margin-bottom:5px;'>Road crossings</div>
         <div style='height:12px;border-radius:2px;margin-bottom:3px;
                     background:linear-gradient(to right,#3b8bc2,#5e4fa2,#8c2d8f,#c31d6b);'></div>
         <div style='display:flex;justify-content:space-between;margin-bottom:5px;'>
           <span>quiet</span><span>busy (AADT)</span></div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:20px;border-top:2.5px solid #555;display:inline-block;'></span>measured</div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:20px;border-top:2.5px dashed #555;display:inline-block;'></span>estimated</div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:15px;height:11px;background:#7d3f1633;border:1px solid #7d3f16aa;
                        display:inline-block;flex:none;'></span>corridor band</div>
         <div style='margin-top:5px;color:#5a6070;font-size:11px;'>
           Thicker = more traffic. Named roads have a dark outline. No puma locations shown.</div>
       </div>"
    )
  ) |>
  fitBounds(-122.17587, 37.01977, -121.69144, 37.28626)
```

## The network, and a distinction that matters

Widening from the single corridor to the whole system of large habitat blocks —
the 29 cores of at least 30 km² — gives a connectivity network: each core linked
to its geometric neighbours, each link routed through the resistance surface. Two
findings come out of it, and holding them apart is the point.

**Coyote Valley is the efficient link, and the traffic barrier — but not the
structural weak point.** The cheapest connections in the network — the strongest
links — run through the south Bay, between the cores on either side of the Coyote
Valley gap. That is where the two halves of the region connect at low cost. Those
south-Bay links are the *efficient* path; the cross-bay alternatives cost far more.
So Coyote Valley matters twice — as the real, protectable connection, and as the
142k traffic barrier astride it — but it is a strong link, not a weak one.

**The structural weak links are the cross-bay spans.** The costliest connections
in the network are the long, high-resistance routes that would join the
Peninsula / Santa Cruz cluster to the East-Bay / Diablo cluster *across the
central urbanized Bay* — spans of 40 to 60 km with no stepping stones between them.
The central Bay effectively splits the system into two subnetworks, joined
efficiently only at the south end, through Coyote Valley. This matches what is
already known: Santa Cruz Mountains pumas are famously isolated, connected to the
Diablo Range mainly via the south Bay.

The traffic pinch and the structural weak point are therefore two different things
on two different axes. Coyote Valley is a traffic-mitigation priority — a real
corridor with a dangerous road across it. The cross-bay gaps are a fragmentation
finding — evidence of what is already disconnected.

**And the cross-bay links are not corridors to build.** This distinction matters
enough to state plainly: a 60 km least-cost path across the developed
bayshore is at the edge of biological plausibility. The network method forces a
route between geometric neighbours whether or not a real one exists, and the
routing dutifully finds the least-bad line across an essentially impassable
barrier. Those longest cross-bay links mean *these clusters are effectively
disconnected*, not *here is a corridor to protect*. Read them as fragmentation
evidence. The south-Bay connection, by contrast, is a real connection worth
protecting.

```{r}
#| label: fig-puma-network
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The puma core-connectivity network over the core habitat it links. Faint green blocks are the core habitat patches. Links between cores are drawn as schematic lines coloured by cost — green links are cheap and strong (the efficient south-Bay connections), red links are costly and weak. The lines show connectivity topology, not exact routes: at a 1 km grain a single-cell path would imply false precision, so the lines say which cores connect and how cheaply, not where a puma walks. The three-to-five costliest links — the structural weak points — are shown instead as red swaths, the honest band form, because those fragile cross-bay connections are the map's real subject. Read the red cross-bay swaths as evidence of disconnection, not as corridors to protect: the method forces a route between geometric neighbours whether or not a real one exists, and a 40–60 km path across the developed bayshore is at the edge of plausibility. No puma locations are shown. Source: Gabriel-graph network (Decision 37, corrected by Decision 38)."

library(sf)
library(leaflet)
library(dplyr)

net <- st_read("../data/processed/lcp_puma_network_3310.gpkg", quiet = TRUE) |>
  st_transform(4326)

# Weak-link swaths — the honest band form for the fragile links (only these 5 were
# swathed by the analysis; the other 33 edges have no swath, so they stay as
# schematic lines). Decision 38: no labelled SC->Diablo chain.
ws <- st_read("../data/processed/lcp_puma_network_weaklinks_swath_3310.gpkg",
              quiet = TRUE) |>
  st_transform(4326)

# Core habitat patches underneath — the cores the links connect.
cp <- st_read("../data/processed/lcp_puma_core_patches_3310.gpkg", quiet = TRUE) |>
  st_transform(4326)

# Cost ordering IS the strong-vs-weak story: cheap = strong, costly = weak.
pal <- colorNumeric(c("#2f8a3e", "#9acd32", "#e8b64c", "#d0691f", "#b3271a"),
                    domain = range(net$cost_distance, na.rm = TRUE))
net$wt <- 4.5 - 1.8 * (net$cost_distance / max(net$cost_distance, na.rm = TRUE))

leaflet(options = leafletOptions(minZoom = 7)) |>
  addTiles(
    urlTemplate = "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}",
    attribution = "Tiles &copy; Esri"
  ) |>
  addTiles(
    urlTemplate = "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Reference/MapServer/tile/{z}/{y}/{x}"
  ) |>
  # core habitat patches — faint context, the cores the links connect
  addPolygons(
    data = cp,
    fillColor = "#5a7d5a", fillOpacity = 0.20,
    color = "#5a7d5a", weight = 0.4, opacity = 0.4,
    label = ~sprintf("Core habitat patch %s (%.0f km²)", patch_id, area_km2)
  ) |>
  # all 38 links as schematic cost-coloured lines (topology, not precise routes)
  addPolylines(
    data = net,
    color = ~pal(cost_distance), weight = ~wt, opacity = 0.85,
    label = ~sprintf("Link %s–%s — cost %.0f, %.0f km",
                     from_patch, to_patch, cost_distance, length_km)
  ) |>
  # top-5 weak-link SWATHS — the honest band form for the fragile cross-bay links
  addPolygons(
    data = ws,
    fillColor = "#b3271a", fillOpacity = 0.15,
    color = "#7a1010", weight = 1.5, opacity = 0.85, dashArray = "5 4",
    label = ~sprintf("Weak link #%d (cross-bay swath) — cost %.0f — disconnection, not a corridor",
                     weak_rank, cost_distance)
  ) |>
  addControl(
    position = "bottomleft",
    html = htmltools::HTML(
      "<div style='background:#fff;padding:8px 10px;border-radius:4px;
                   box-shadow:0 1px 4px rgba(0,0,0,0.2);font-size:12px;
                   line-height:1.4;text-align:left;max-width:220px;'>
         <div style='font-weight:600;margin-bottom:5px;'>Core-to-core links</div>
         <div style='height:12px;border-radius:2px;margin-bottom:3px;
                     background:linear-gradient(to right,#2f8a3e,#9acd32,#e8b64c,#d0691f,#b3271a);'></div>
         <div style='display:flex;justify-content:space-between;margin-bottom:5px;'>
           <span>cheap · strong</span><span>costly · weak</span></div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:15px;height:11px;background:#b3271a44;border:1px dashed #7a1010;
                        display:inline-block;flex:none;'></span>weak-link swaths (cross-bay)</div>
         <div style='display:flex;align-items:center;gap:7px;'>
           <span style='width:15px;height:11px;background:#5a7d5a44;border:1px solid #5a7d5a;
                        display:inline-block;flex:none;'></span>core habitat</div>
         <div style='margin-top:5px;color:#5a6070;font-size:11px;'>
           Lines are connectivity topology, not exact routes. Red cross-bay swaths = disconnection evidence, not corridors. No puma locations shown.</div>
       </div>"
    )
  ) |>
  fitBounds(-123.53354, 36.85065, -121.20830, 38.86399)
```

## What this section is, and is not

These corridors are hypotheses about movement drawn from a landscape-cost model,
not tracks from collared animals. No puma was followed to draw them; there is no
telemetry here to confirm them. Shifting the model's weights within a plausible
range moves the corridor by up to about 1.5 km, so the routes are stable in
location but not exact to the metre. They are the best structural estimate of where
connectivity runs — a well-founded hypothesis, and the right basis for asking where
to look and where to protect, but a hypothesis.

Every puma layer here is shown as generalised geometry — bands and routes, never
points. That is deliberate and required: precise puma locations are a poaching and
disturbance risk, so the project never publishes them. Nothing in this section is
fine enough to reverse-engineer a den, a kill site, or a camera. The corridors are
about where an animal *could* move, at the scale of the landscape, not about where
any individual animal *is*.
