---
title: "What was recorded, and what is there"
subtitle: "Q5 — effort, and the two species read differently"
---

## The question behind the other four

Every map on this site is built from records that people made. People do not
look everywhere equally. They look near roads, near trails, near cities, and in
the months of survey drives and community bio-blitzes. So a map of *where the
records are* is partly a map of *where the observers were*. This is the effort
problem, and Q5 asks it directly: how much of the pattern is the animal, and how
much is the observer?

This page does not add a new dataset. It reads the puma and the bobcat results
against each other, because the two species answer the effort question in
opposite ways — and the contrast is itself the finding.

::: {.callout-note title="Read this after the Bobcat and Puma pages"}
Q5 is a cross-read. It uses the bobcat occupancy result (the Bobcat page) and the
puma corridor result (the Puma page). If you have not seen those, the short
version is below each heading.
:::

## Two species, two relationships

A descriptive surface — a kernel-density map of records, or a hot-spot map —
carries the effort signal directly, because it is built from the raw records. The
two analytical tracks then do different things with that signal.

- **Bobcat — the model pulls away from effort.** The occupancy model has a
  separate detection sub-model. It estimates *effort* and *presence* as two
  things, then reports presence with the effort taken out. So the modelled result
  is expected to move away from the raw record density.
- **Puma — the model never sees effort at all.** The corridor is built from the
  landscape: land cover, human modification, slope, and traffic (see the Methods
  page and Decision 26). There is no record-density term anywhere in it. So if
  the corridor lands on the observed record density, that agreement was not put
  there by construction.

The result is a divergence for one species and a convergence for the other, and
both are honest for the same reason.

### Bobcat: modelled occupancy DIVERGES from the record density

The bobcat occupancy surface and the raw kernel-density surface do not match. The
correlation between the two is near zero (r = 0.075).

That is the correct outcome, not a failure. The raw density is effort-shaped —
high where people surveyed. The occupancy model corrects for that, because its
detection sub-model absorbs the survey effort and reports the presence pattern
underneath. The two surfaces disagree *because the model is doing its job.* A high
agreement here would have been a warning sign that the model had simply re-drawn
the effort map.

### Puma: modelled corridor CONVERGES with the record density

The puma corridor is the opposite case, and it also reverses what we expected
before we ran it.

The pre-analysis guess was **divergence** — that the corridor, and especially the
Coyote Valley pinch, would thread through *low-record* gaps, pumas slipping
through a developed squeeze unseen. The data says the opposite:

- The corridor runs through **above-median** record density along its whole
  length. The least-cost path sits at the **77th percentile** of the density
  surface (median across the route), and **none** of it falls below the median.
- The **Coyote Valley pinch is at the 83rd percentile** — high record density,
  not the predicted low.
- Of the record hot-spots the corridor crosses, **six of seven are effort-checked
  as trusted** — real record concentration, not survey artifact.

Two methods that share no effort term — a structure-driven least-cost corridor and
an effort-shaped record surface — agree on where the corridor is. Because the
corridor model never saw the records, that agreement is evidence the puma pattern
is real terrain-driven signal, not an effort shadow.

## The convergence, on the map

The surface below is the puma record density (kernel density, published at the
1 km grain the sensitive-data policy requires — see the note at the foot of this
page). On top of it are the two-tier corridor swath and the core habitat patches.
Read it for one thing: **the corridor band sits on the warmer parts of the
surface, not the gaps.**

```{r}
#| label: fig-puma-convergence
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Puma record density (kernel-density surface, 1 km) with the two-tier movement corridor and core habitat patches. The corridor band runs through above-median density along its length, including at the Coyote Valley pinch. Density is a smoothed record surface and is effort-shaped; it is read here only as a cross-check on the structure-built corridor, never as a distribution result on its own."

library(terra)
library(sf)
library(leaflet)
library(htmltools)

# --- paths: page-relative from site/ (NEVER here::here) ---
kde_path    <- "../data/processed/kde_puma_current_1km_3310.tif"
swath_path  <- "../data/processed/lcp_puma_scmtns_to_diablo_swath_3310.gpkg"
patch_path  <- "../data/processed/lcp_puma_core_patches_3310.gpkg"

# --- read ---
kde_r   <- terra::rast(kde_path)
swath_sf <- sf::st_read(swath_path, quiet = TRUE)
patch_sf <- sf::st_read(patch_path, quiet = TRUE)

# --- reproject to WGS84 for leaflet ---
kde_wgs   <- terra::project(kde_r, "EPSG:4326", method = "bilinear")
swath_wgs <- sf::st_transform(swath_sf, 4326)
patch_wgs <- sf::st_transform(patch_sf, 4326)

# --- sqrt stretch (matches bobcat KDE page); NAs stay transparent ---
v <- terra::values(kde_wgs)
v <- v[is.finite(v) & v > 0]
cap <- as.numeric(stats::quantile(v, 0.99, na.rm = TRUE))
kde_stretch <- terra::clamp(kde_wgs, lower = 0, upper = cap, values = TRUE)
kde_stretch <- terra::app(kde_stretch, fun = function(x) sqrt(x))

rng <- terra::minmax(kde_stretch)
dom <- c(as.numeric(rng[1]), as.numeric(rng[2]))

# tiger-green ramp, consistent with the bobcat KDE map; NA -> transparent
ramp_cols <- c("#f7fcf5", "#c7e9c0", "#74c476", "#31a354", "#006d2c")
pal <- leaflet::colorNumeric(ramp_cols, domain = dom, na.color = "transparent")

# --- corridor swath tiers ---
swath_core    <- swath_wgs[swath_wgs$tier == "core", ]
swath_context <- swath_wgs[swath_wgs$tier == "context", ]

# --- endpoints from linkage_role (no hardcoded ids) ---
patch_end <- patch_wgs[!is.na(patch_wgs$linkage_role), ]

# --- custom left-aligned legend (addLegend cannot mix a ramp + swatch styles) ---
legend_html <- paste0(
  "<div style='background:rgba(255,255,255,0.9);padding:8px 10px;",
  "border-radius:4px;font:12px/1.4 sans-serif;text-align:left;'>",
  "<div style='font-weight:600;margin-bottom:4px;'>Puma cross-read</div>",
  "<div><span style='display:inline-block;width:14px;height:10px;",
  "background:linear-gradient(90deg,#f7fcf5,#31a354,#006d2c);",
  "vertical-align:middle;margin-right:6px;'></span>Record density (low &rarr; high)</div>",
  "<div><span style='display:inline-block;width:14px;height:10px;",
  "background:#d94801;opacity:0.55;vertical-align:middle;margin-right:6px;'></span>",
  "Corridor core band (&le; q2%)</div>",
  "<div><span style='display:inline-block;width:14px;height:10px;",
  "background:#fdae6b;opacity:0.45;vertical-align:middle;margin-right:6px;'></span>",
  "Corridor context band (&le; q5%)</div>",
  "<div><span style='display:inline-block;width:12px;height:12px;",
  "border:1.5px solid #555;background:rgba(0,0,0,0);",
  "vertical-align:middle;margin-right:6px;'></span>Core habitat patch</div>",
  "</div>"
)

leaflet(options = leafletOptions(minZoom = 8)) |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}"),
    attribution = "Tiles &copy; Esri") |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Reference/MapServer/tile/{z}/{y}/{x}")) |>
  # density surface
  addRasterImage(kde_stretch, colors = pal, opacity = 0.75, project = FALSE) |>
  # core patches, faint (context, not the point of the map)
  addPolygons(data = patch_wgs, fill = FALSE, color = "#555555",
              weight = 0.7, opacity = 0.5) |>
  # corridor context band (wider, lighter), then core band on top
  addPolygons(data = swath_context, stroke = FALSE,
              fillColor = "#fdae6b", fillOpacity = 0.45) |>
  addPolygons(data = swath_core, stroke = FALSE,
              fillColor = "#d94801", fillOpacity = 0.55) |>
  # endpoints, bold outline
  addPolygons(data = patch_end, fill = FALSE, color = "#222222",
              weight = 2.2, opacity = 0.9) |>
  addControl(html = HTML(legend_html), position = "bottomleft") |>
  # fitBounds MUST be last, or the auto-fit overrides it
  fitBounds(-123.53354, 36.85065, -121.20830, 38.86399)
```

The density surface is smoothed with a wide (5 km) bandwidth, set from a puma
home-range prior. That width is why the map shows broad warm zones, not sharp
points — and it is one reason the percentile figures stay in the text and not on
the legend: a smoothed surface cannot honestly show a single-cell pinch value.

## The record hot-spots, and how many are just effort

The density surface above is smooth. A hot-spot map is sharper: it marks the
open-space units where puma records cluster **more than their neighbours**, tested
for significance. But a record hot-spot has two possible causes — real puma
presence, or a place people surveyed hard. The Q5 effort check separates them:

- **Trusted** — the unit is a record hot-spot **and not** an effort hot-spot. Real
  concentration.
- **Suspect** — the unit is a record hot-spot **and also** an effort hot-spot. It
  may just be where people looked.

The map below draws trusted units as solid fill and suspect units as hollow,
dashed outlines. Read the split, not just the count: for puma, **about half of all
hot units are effort-suspect** — far more than for bobcat. That is the effort
problem made visible on a single map.

```{r}
#| label: fig-puma-hotspots
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "!expr caption_txt"

# libraries already attached by fig-puma-convergence (terra, sf, leaflet, htmltools)

hot_path <- "../data/processed/hot_puma_gistar_unit_3310.gpkg"
hot_sf   <- sf::st_read(hot_path, quiet = TRUE)

# --- hot units only, split by Q5 effort flag ---
hot_all <- hot_sf[!is.na(hot_sf$hotspot) & hot_sf$hotspot == "hot", ]

# TRUSTED = occ-hot but NOT effort-hot; SUSPECT = occ-hot AND effort-hot
is_trusted <- grepl("TRUSTED", hot_all$q5_flag)
is_suspect <- grepl("SUSPECT", hot_all$q5_flag)

# --- SMALL-n COARSEN GUARD (sensitive-data policy §3) ---
# A hot unit resting on 1-2 precise records can reverse-narrow toward a location.
# Coarsen those: draw the polygon, but NO popup / NO label. Count only.
SMALL_N <- 3L
is_smalln <- !is.na(hot_all$n_occ) & hot_all$n_occ < SMALL_N
n_coarsened <- sum(is_smalln)

trusted_sf <- hot_all[is_trusted, ]
suspect_sf <- hot_all[is_suspect, ]

# popups ONLY for units at/above the small-n floor; coarsened units get no label
mk_popup <- function(df) {
  ifelse(!is.na(df$n_occ) & df$n_occ >= SMALL_N,
         paste0("<b>", htmltools::htmlEscape(df$unit_name), "</b>"),
         NA_character_)
}

trusted_wgs <- sf::st_transform(trusted_sf, 4326)
suspect_wgs <- sf::st_transform(suspect_sf, 4326)
trusted_wgs$popup <- mk_popup(trusted_sf)
suspect_wgs$popup <- mk_popup(suspect_sf)

# --- computed caption (numbers come from the data, never typed) ---
caption_txt <- sprintf(
  paste0("Puma record hot-spots (Getis-Ord Gi*, FDR-corrected) by open-space unit. ",
         "Solid = trusted (%d units: record hot-spot, not an effort hot-spot); ",
         "hollow dashed = effort-suspect (%d units: also an observer-effort hot-spot). ",
         "About half of all puma hot units are effort-suspect \u2014 far more than bobcat ",
         "\u2014 which is the effort problem made visible. %s ",
         "Hot-spots are record-driven and effort-shaped; they are not a puma distribution."),
  nrow(trusted_wgs), nrow(suspect_wgs),
  if (n_coarsened > 0)
    sprintf("%d hot unit(s) resting on fewer than %d precise records are drawn without a label, per the sensitive-data policy.",
            n_coarsened, SMALL_N)
  else
    "No hot unit rested on fewer than three precise records."
)

# custom left-aligned legend (solid vs hollow-dashed swatches)
legend_html2 <- paste0(
  "<div style='background:rgba(255,255,255,0.9);padding:8px 10px;",
  "border-radius:4px;font:12px/1.4 sans-serif;text-align:left;'>",
  "<div style='font-weight:600;margin-bottom:4px;'>Puma record hot-spots (Gi*)</div>",
  "<div><span style='display:inline-block;width:14px;height:10px;",
  "background:#cb181d;opacity:0.7;vertical-align:middle;margin-right:6px;'></span>",
  "Trusted (not effort-driven)</div>",
  "<div><span style='display:inline-block;width:14px;height:10px;",
  "background:rgba(0,0,0,0);border:1.5px dashed #cb181d;",
  "vertical-align:middle;margin-right:6px;'></span>",
  "Effort-suspect (where people looked)</div>",
  "</div>"
)

leaflet(options = leafletOptions(minZoom = 8)) |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}"),
    attribution = "Tiles &copy; Esri") |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Reference/MapServer/tile/{z}/{y}/{x}")) |>
  # suspect: hollow, dashed red outline, no fill
  addPolygons(data = suspect_wgs, fill = FALSE,
              color = "#cb181d", weight = 1.5, opacity = 0.9,
              dashArray = "5", popup = ~popup) |>
  # trusted: solid red fill on top
  addPolygons(data = trusted_wgs, stroke = TRUE, color = "#cb181d",
              weight = 0.6, opacity = 0.8,
              fillColor = "#cb181d", fillOpacity = 0.7,
              popup = ~popup) |>
  addControl(html = HTML(legend_html2), position = "bottomleft") |>
  # fitBounds MUST be last
  fitBounds(-123.53354, 36.85065, -121.20830, 38.86399)
```

The hot-spot map and the corridor tell the same story from two directions. Where
the corridor crosses record hot-spots, most of those units are **trusted** — real
presence, not effort. That is the part of the convergence that is not circular. But
across the whole region, the large share of **suspect** units is the reminder Q5
exists to give: a red unit is not proof of pumas until the effort check clears it.

## What this convergence is, and what it is not

The agreement is real, but it must be read at its true strength — corroboration,
not proof.

- **It is corroboration.** Two methods with no shared effort term land in the same
  places, and most of the record hot-spots on the corridor are effort-checked as
  trusted. The puma pattern behaves like real terrain-driven signal.
- **It is only weak-to-moderate, and partly structural.** The resistance surface
  and the density surface share a land-cover and human-modification foundation. So
  part of the agreement is two views of the *same* landscape gradient, not two
  independent confirmations. A corridor built to follow good habitat, checked
  against records that also concentrate in good habitat, is *expected* to agree —
  the ecology literature says as much.
- **It is not validation.** Strong, independent validation needs movement data —
  GPS collars showing pumas actually using the route. This project has occurrence
  records only. The corridor is a plausible, record-corroborated hypothesis, not a
  confirmed path.

Two things keep the convergence meaningful in spite of the shared foundation. The
trusted hot-spots are record concentration that is **not** effort-driven — real
presence on the corridor. And the corridor carries a **road-and-traffic term the
density surface does not have** — so the two are not the same construct, and
agreement across that difference is more than shared-terrain overlap.

::: {.callout-important title="Why we lead with the band, not the line"}
The corridor is shown as a two-tier **swath**, not a single line. Against
simulated true movement, the precise least-cost centre-line is the least accurate
connectivity form, while a swath overlaps true movement far better. The line is a
tidy artifact of the algorithm; the band is the more honest product. This is why
the pinch percentile (83rd) is the *cell* value near the route, not an on-line
reading — at 1 km grain the centre-line passes about 1.6 km from the exact pinch
cell.
:::

## The shared caveat, stated once

Both species inherit the same ceiling, and it is the point of Q5: **descriptive
record surfaces are effort-shaped.** A kernel-density map or a raw hot-spot map
shows where records concentrate, and records concentrate where people looked. The
two tracks handle this differently — the bobcat occupancy model removes the effort
signal and so *diverges* from the raw surface; the puma corridor never contains an
effort signal and so its *convergence* with the raw surface is informative rather
than circular — but neither track escapes the fact that the underlying records
were made by people, unevenly. Every descriptive map on this site should be read
with that in mind.

::: {.callout-note title="Sensitive-data note (puma)"}
Precise puma locations are withheld. iNaturalist does not obscure mountain lions
in California, so this project holds precise records that could expose real
animals if published as points. The density surface on this page is published only
at a 1 km grain, which cannot be reversed to a location, and it clears the
project's publish gate at that grain. No puma points, den sites, or kill sites
appear anywhere on this site.
:::
