---
title: "What was recorded, and what is there?"
subtitle: "Effort, and the two species"
---

## The question behind the other four

Every map on this site is built from records that people made. People do not
look everywhere equally. They look near roads, near trails, near cities, and in
the months of survey drives and community bio-blitzes. So a map of *where the
records are* is partly a map of *where the observers were*. This is the effort
problem: how much of the pattern is the animal, and how
much is the observer?

## Two species, two relationships

A descriptive surface — a kernel-density map of records, or a hot-spot map —
carries the effort directly, because it is built from the raw records. The
two analytical tracks then do different things with that.

- **Bobcat — the model pulls away from effort.** The occupancy model has a
  separate detection sub-model. It estimates *effort* and *presence* as two
  things, then reports presence with the effort taken out. So the modelled result
  is expected to move away from the raw record density.
- **Puma — the model never sees effort at all.** The corridor is built from the
  landscape: land cover, human modification, slope, and traffic. 
  There is no record-density term anywhere in it. So if
  the corridor lands on the observed record density, that agreement was not put
  there by construction.

The result is a divergence for one species and a convergence for the other.

### Bobcat: modeled occupancy DIVERGES from the record density

The bobcat occupancy surface and the raw kernel-density surface do not match. The
correlation between the two is near zero (r = 0.075).

The raw density is effort-shaped — high where people surveyed.
The occupancy model corrects for that, because its
detection sub-model absorbs the survey effort and reports the presence.
A high agreement here would have been a warning sign that the model had simply re-drawn
the effort map.

### Puma: modeled corridor CONVERGES with the record density

The puma corridor is the opposite case:

- The corridor runs through **above-median** record density along its whole
  length. The least-cost path sits at the **77th percentile** of the density
  surface (median across the route), and **none** of it falls below the median.
- The **Coyote Valley pinch is at the 83rd percentile** — high record density.
- Of the record hot-spots the corridor crosses, **six of seven are effort-checked
  as trusted**.

## Kernel Density and Hot spots

The surface below is the puma record density (kernel density). On top of it are 
the two-tier corridor swath and the core habitat patches.
**The corridor band sits on the warmer parts of the
surface.**

```{r}
#| label: fig-puma-convergence
#| column: page
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Puma record density (kernel-density surface, 1 km) with the two-tier movement corridor and core habitat patches. The density surface is smoothed with a wide (5 km) bandwidth. The corridor band runs through above-median density along its length, including at the Coyote Valley pinch. Density is a smoothed record surface and is effort-shaped."

library(terra)
library(sf)
library(leaflet)
library(htmltools)

# --- paths: page-relative from site/ (NEVER here::here) ---
kde_path    <- "../data/processed/kde_puma_current_1km_3310.tif"
swath_path  <- "../data/processed/lcp_puma_scmtns_to_diablo_swath_3310.gpkg"
patch_path  <- "../data/processed/lcp_puma_core_patches_3310.gpkg"

# --- read ---
kde_r   <- terra::rast(kde_path)
swath_sf <- sf::st_read(swath_path, quiet = TRUE)
patch_sf <- sf::st_read(patch_path, quiet = TRUE)

# --- reproject to WGS84 for leaflet ---
kde_wgs   <- terra::project(kde_r, "EPSG:4326", method = "bilinear")
swath_wgs <- sf::st_transform(swath_sf, 4326)
patch_wgs <- sf::st_transform(patch_sf, 4326)

# --- sqrt stretch (matches bobcat KDE page); NAs stay transparent ---
v <- terra::values(kde_wgs)
v <- v[is.finite(v) & v > 0]
cap <- as.numeric(stats::quantile(v, 0.99, na.rm = TRUE))
kde_stretch <- terra::clamp(kde_wgs, lower = 0, upper = cap, values = TRUE)
kde_stretch <- terra::app(kde_stretch, fun = function(x) sqrt(x))

rng <- terra::minmax(kde_stretch)
dom <- c(as.numeric(rng[1]), as.numeric(rng[2]))

# tiger-green ramp, consistent with the bobcat KDE map; NA -> transparent
ramp_cols <- c("#f7fcf5", "#c7e9c0", "#74c476", "#31a354", "#006d2c")
pal <- leaflet::colorNumeric(ramp_cols, domain = dom, na.color = "transparent")

# --- corridor swath tiers ---
swath_core    <- swath_wgs[swath_wgs$tier == "core", ]
swath_context <- swath_wgs[swath_wgs$tier == "context", ]

# --- endpoints from linkage_role (no hardcoded ids) ---
patch_end <- patch_wgs[!is.na(patch_wgs$linkage_role), ]

# --- custom left-aligned legend (addLegend cannot mix a ramp + swatch styles) ---
legend_html <- paste0(
  "<div style='background:rgba(255,255,255,0.9);padding:8px 10px;",
  "border-radius:4px;font:12px/1.4 sans-serif;text-align:left;'>",
  "<div style='font-weight:600;margin-bottom:4px;'>Puma cross-read</div>",
  "<div><span style='display:inline-block;width:14px;height:10px;",
  "background:linear-gradient(90deg,#f7fcf5,#31a354,#006d2c);",
  "vertical-align:middle;margin-right:6px;'></span>Record density (low &rarr; high)</div>",
  "<div><span style='display:inline-block;width:14px;height:10px;",
  "background:#d94801;opacity:0.55;vertical-align:middle;margin-right:6px;'></span>",
  "Corridor core band (&le; q2%)</div>",
  "<div><span style='display:inline-block;width:14px;height:10px;",
  "background:#fdae6b;opacity:0.45;vertical-align:middle;margin-right:6px;'></span>",
  "Corridor context band (&le; q5%)</div>",
  "<div><span style='display:inline-block;width:12px;height:12px;",
  "border:1.5px solid #555;background:rgba(0,0,0,0);",
  "vertical-align:middle;margin-right:6px;'></span>Core habitat patch</div>",
  "</div>"
)

leaflet(options = leafletOptions(minZoom = 8)) |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}"),
    attribution = "Tiles &copy; Esri") |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Reference/MapServer/tile/{z}/{y}/{x}")) |>
  # density surface
  addRasterImage(kde_stretch, colors = pal, opacity = 0.75, project = FALSE) |>
  # core patches, faint (context, not the point of the map)
  addPolygons(data = patch_wgs, fill = FALSE, color = "#555555",
              weight = 0.7, opacity = 0.5) |>
  # corridor context band (wider, lighter), then core band on top
  addPolygons(data = swath_context, stroke = FALSE,
              fillColor = "#fdae6b", fillOpacity = 0.45) |>
  addPolygons(data = swath_core, stroke = FALSE,
              fillColor = "#d94801", fillOpacity = 0.55) |>
  # endpoints, bold outline
  addPolygons(data = patch_end, fill = FALSE, color = "#222222",
              weight = 2.2, opacity = 0.9) |>
  addControl(html = HTML(legend_html), position = "bottomleft") |>
  # fitBounds MUST be last, or the auto-fit overrides it
  fitBounds(-123.53354, 36.85065, -121.20830, 38.86399)
```

## The record hot-spots

The density surface above is smooth. A hot-spot map is sharper: it marks the
open-space units where puma records cluster **more than their neighbours**, tested
for significance. But a record hot-spot has two possible causes — real puma
presence, or a place people surveyed hard:

- **Trusted** — the unit is a record hot-spot **and not** an effort hot-spot. Real
  concentration.
- **Suspect** — the unit is a record hot-spot **and also** an effort hot-spot. It
  may just be where people looked.

The map below draws trusted units as solid fill and suspect units as hollow,
dashed outlines. For the puma analysis, **about half of all
hot units are effort-suspect** — far more than for bobcat.

```{r}
#| label: fig-puma-hotspots
#| column: page
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Getis-Ord Gi* hot-spot units for puma records, FDR-corrected. Solid red is a TRUSTED hotspot; hollow dashed red is a hotspot that also coincides with a recording-effort hotspot, flagged effort-SUSPECT."

# libraries already attached by fig-puma-convergence (terra, sf, leaflet, htmltools)

hot_path <- "../data/processed/hot_puma_gistar_unit_3310.gpkg"
hot_sf   <- sf::st_read(hot_path, quiet = TRUE)

# --- hot units only, split by Q5 effort flag ---
hot_all <- hot_sf[!is.na(hot_sf$hotspot) & hot_sf$hotspot == "hot", ]

# TRUSTED = occ-hot but NOT effort-hot; SUSPECT = occ-hot AND effort-hot
trusted_sf <- hot_all[grepl("TRUSTED", hot_all$q5_flag), ]
suspect_sf <- hot_all[grepl("SUSPECT", hot_all$q5_flag), ]

# --- SMALL-n COARSEN GUARD (sensitive-data policy §3) ---
# A hot unit resting on 1-2 precise records can reverse-narrow toward a location.
# Coarsen those: draw the polygon, but give it NO hover label. Count only.
SMALL_N     <- 3L
n_coarsened <- sum(!is.na(hot_all$n_occ) & hot_all$n_occ < SMALL_N)

trusted_wgs <- sf::st_transform(trusted_sf, 4326)
suspect_wgs <- sf::st_transform(suspect_sf, 4326)

# --- colours: match bobcat Gi* map exactly ---
col_hot     <- "#B30000"
stroke_hot  <- "#7a0000"

# --- hover label (bobcat lab_common convention); NULL for small-n units ---
lab_common <- function(d, verdict = NULL) {
  keep <- !is.na(d$n_occ) & d$n_occ >= SMALL_N   # small-n get no detail
  base <- sprintf("%s (unit %s)<br>Gi* z = %.2f, FDR p = %.3f",
                  d$unit_name, d$unit_id, d$gistar_z, d$gistar_p_fdr)
  if (!is.null(verdict)) base <- paste0(base, "<br><b>", verdict, "</b>")
  base[!keep] <- ""                              # blank label = no tooltip shown
  lapply(base, htmltools::HTML)
}

# --- legend: mirror the bobcat swatch style (%s22 hex-alpha for suspect fill) ---
legend_html2 <- htmltools::HTML(sprintf(
  "<div style='background:#fff;padding:8px 10px;border-radius:4px;
               box-shadow:0 1px 4px rgba(0,0,0,0.2);font-size:12px;
               line-height:1.5;text-align:left;'>
     <div style='font-weight:600;margin-bottom:5px;'>Puma record hot-spots (Gi*, FDR-corrected)</div>
     <div style='display:flex;align-items:center;gap:7px;'>
       <span style='width:15px;height:13px;background:%s;border:1px solid #7a0000;
                    display:inline-block;flex:none;'></span>Hotspot &mdash; trusted</div>
     <div style='display:flex;align-items:center;gap:7px;'>
       <span style='width:15px;height:13px;background:%s22;border:1.6px dashed %s;
                    display:inline-block;flex:none;'></span>Hotspot &mdash; effort-suspect</div>
   </div>",
  col_hot, col_hot, col_hot
))

leaflet(options = leafletOptions(minZoom = 8)) |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Base/MapServer/tile/{z}/{y}/{x}"),
    attribution = "Tiles &copy; Esri") |>
  addTiles(urlTemplate = paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_Light_Gray_Reference/MapServer/tile/{z}/{y}/{x}")) |>
  # suspect hot — faint red fill + dashed outline (bobcat convention, NOT fully hollow)
  addPolygons(data = suspect_wgs,
              fillColor = col_hot, fillOpacity = 0.12,
              color = col_hot, weight = 1.6, opacity = 0.95, dashArray = "4 3",
              label = lab_common(suspect_sf,
                "EFFORT-SUSPECT hotspot (coincides with an effort hotspot)")) |>
  # trusted hot — solid red on top
  addPolygons(data = trusted_wgs,
              fillColor = col_hot, fillOpacity = 0.85,
              color = stroke_hot, weight = 1.0, opacity = 0.9,
              label = lab_common(trusted_sf,
                "TRUSTED hotspot (not explained by effort)")) |>
  addControl(html = legend_html2, position = "bottomleft") |>
  # fitBounds MUST be last
  fitBounds(-123.53354, 36.85065, -121.20830, 38.86399)
```

::: {.aside}
Of the puma hot units, the effort cross-read trusts fewer than half; the rest
coincide with recording-effort hotspots and are flagged suspect. Any hot unit
resting on fewer than three precise records is drawn without a hover label.
:::

::: {.callout-note title="Sensitive-data note (puma)"}
Precise puma locations are withheld. iNaturalist does not obscure mountain lions
in California, so this project does not share precise records that could expose real
animals if published as points. The density surface on this page is published only
at a 1 km grain, which cannot be reversed to a location. No puma points, den sites, or kill sites
appear anywhere on this site.
:::
